Supporting information: Tail associations and their impact on extinction risk

### 2019-11-05 MainText.Rmd
edits

The stochastic quantities $s_e(t)$ and $\tilde{f}_a(t)$ were generated 
to have the same overall association (Spearman's $\rho$ equal to $0.875$,
up to sampling variance), but different degrees and types of 
asymmetry of tail association in different simulations, as follows.
For a simulation of length $T$, we used methods that can 
generate $T$ independent pairs $(u(t),v(t))$ for $t=1,\ldots,T$ with
$\rho=0.875$ (up to sampling variance), but with either extreme left-,
moderate left-, moderate right-, or extreme right-tail association 
(e.g., Fig. \ref{fig_intro_cop}). The methods we used generated
$(u(t),v(t))$ such that $u(t)$ was uniformly distributed on $(0,1)$,
and likewise for $v(t)$. We then applied a monotonic transformation
to the $u(t)$ so that the resulting numbers were distributed according to
a beta distribution with parameters
$\alpha=3$ and $\beta=2$, 
and we used those values for the $s_e(t)$. We also applied a (different)
monotonic transformation to the $v(t)$ so that the resulting numbers were
distributed according to a gamma distribution with parameters $k=2$
and $\theta=1$, and we used those values for the $\tilde{f}_a(t)$. Because
the $u(t)$ and $v(t)$ were uniformly distributed, the inverse 
cummulative distribution functions of the desired beta and
gamma distributions were the transformations. Because these
were monotonic transformations, the Spearman correlation of 
$s_e(t)$ and $\tilde{f}_a(t)$ was the same as that of 
$u(t)$ and $v(t)$. The quantities $(u(t),v(t))$ for $t=1,\ldots,T$ were
generated as described in Appendix \ref{SM-makenoise}.

@@

We simulated our metapopulation models using environmental noises $\epsilon_i(t)$ (Table \ref{tab_model_summary}), 
independent through time, that were generated to have the same overall associations 
between all patches and in all simulations (Spearman's $\rho$ again `r srho_common`, up to sampling variance), 
but with different degrees and types of asymmetry of tail associations 
between patches in different simulations, as follows. For a simulation of length $T$,
we used methods that can generate $T$ independent $n$-tuples $(u_1(t),\ldots,u_n(t))$
for $t=1,\ldots,T$ with $\rho=$ `r srho_common` between $u_i(t)$ and $u_j(t)$ for all $i \neq j$
(up to sampling variance), but with either extreme left-, moderate left-, moderate right-,
or extreme right-tail association between all pairs of locations (as in Fig. \ref{fig_intro_cop}). 
Pairs $u_i(t)$ and $u_j(t)$ looked like Fig. \ref{fig_intro_cop} and were structured, statistically,
in the same way as the stochastic quantities $u(t)$ and $v(t)$ used for the non-spatial 
simulations described above. For any $i$, $u_i(t)$ was again uniformly distributed on 
the interval $(0,1)$. To obtain the values $\epsilon_i(t)$, the $u_i(t)$ were 
transformed so that the resulting numbers were
normally distributed with mean $0$ and standard deviation $\sigma$ (Table \ref{tab_model_summary}), 
for all $i$. This was 
accomplished with the inverse cummulative distribution function of the normal distribution
with mean $0$ and standard deviation $\sigma$. Because this was a monotonic transformation,
the Spearman correlation between $\epsilon_i(t)$ and $\epsilon_j(t)$ was the same 
as that between $u_i(t)$ and $u_j(t)$; likewise partial Spearman correlations were the same. 
The quantities $(u_1(t),\ldots,u_n(t))$
for $t=1,\ldots,T$ were generated as described in Appendix \ref{SM-makenoise}.

@@

Because the deterministic, one-patch versions of our models showed undercompensatory
dynamics for $r<r_c$ (and therefore for $r$ on the left end of the range $(r_{min},r_{bf})$),
and showed overcompensatory dynamics for $r>r_c$, Figs \ref{fig_extrisk_allmodel_localdisp}
and \ref{SM-fig_extrisk_allmodel_globaldisp} suggest the hypothesis that the
opposite effects of tail association on extinction risk correspond to the domains
in which models exhibit under- versus overcompensatory dynamics. This was 
approximately, but not exactly, true. Careful inspection of Figs 
\ref{SM-fig_risk_local_disp_verhulst_r_vary_scl_0.6} - \ref{SM-fig_risk_local_disp_abrewer_r_vary_scl_0.6} and Figs \ref{SM-fig_risk_global_disp_verhulst_r_vary_scl_0.6} - \ref{SM-fig_risk_global_disp_abrewer_r_vary_scl_0.6}  
and the $r_c$ values of 
Table \ref{tab_model_summary} reveals that, for the Velhurst and Austin-Brewer models,
the switch occurs at a value of $r$ less than $r_c$ for these models. For the example
case of $d=0$, plots of extinction risk versus $r$ (Fig. \ref{fig_extrisk_d0}) seem to indicate that,
for the Ricker, Hassell, and Pennycuick models, left-tail association accentuates and
left-tail association mitigates extinction risk for undercompensatory dynamics ($r<r_c$),
whereas the reverse occurs for overcompensatory dynamics ($r>r_c$). Fig. \ref{fig_extrisk_d0}
reveals a similar reversal for the Maynard Smith, Velhurst and Austin-Brewer models; but 
it occurs at a value of $r$ close to but distinguishable from $r_c$.

@@

<!--linking back to ECEs-->
Extensive evidence now exists that ECEs have become more frequent and more intense
over the last several decades, and substantial progress has recently been made so that
some event categories and even specific events can now be attributed to
human-induced climate change [@ummenhofer2017extreme]. We here provide a fuller 
explanation of how such changes may modify the asymmetry of tail association of 
weather variables measured in different places, and thereby influence metapopulation
extinction risk through the mechansms we explored in this paper. 
If $X_i(t)$ denotes the value of an environmental variable (e.g., a summer maximum 
temperature, or winter minimum temperature, spring rainfall, etc.) 
in location $i$ in year $t$,
@ummenhofer2017extreme provide a diagram of how ECEs may be increasing in frequency 
and intensity due to changes in the distribution of $X_i(t)$ such as changes
in the mean, variance, or skewness of this distribution. Although changes in 
the correlation between $X_i(t)$ and $X_j(t)$ are increasingly documented, 
and may also be a consequence of climate change (e.g., @Black2018),
we are unaware of any studies examining the extent to which tail associations
between $X_i(t)$ and $X_j(t)$, and asymmetries of tail association, may be 
changing. Documented changes in correlation between $X_i(t)$ and $X_j(t)$
may be due partly or wholly to changes in tail associations. If ECEs are becoming
more spatially extensive, as seems likely given that they are becoming
more intense [@ummenhofer2017extreme], then the conditional probability that both of
$X_i(t)$ and $X_j(t)$ are influenced by an ECE given that one of them is so
influenced should be increasing. This should contribute to increased association
of the two variables in one of their tails (which tail depends on the type of 
ECE and what the variable $X$ is). A study of changes through time in 
tail associations of weather variables would be straightforward and 
potentially important given the mechanisms we have explored.
<!--***DAN: Shyamolina, another aspect of an exploration you may do on the 
the geography of tail association may be to look at how tail association 
may be changing. This relates to ECEs, so may help better justify the work
to a funder. You had proposed to look at geography of tail association in 
NDVI, and that makes good sense. You could also look at how tail association
is changing in NDVI (the records are about 35 years long, so you could just
about do this). Even better (because data are better), you could get weather
station records (lots of these can be downloaded, over 1000 stations
have been operating for over a century continuously in North America) and 
look at how tail associations are changing in those data.-->

<!--copulas and rank methods-->
It is known that the distributions $X_i(t)$ are changing, both in
mean and variance, and possibly in other ways, depending on what
environmental variable $X$ represents.
Studying associations between the values of an environmental variable in different
locations, $X_i(t)$ and $X_j(t)$, using standard methods such as Pearson correlation
may be complicated by these changes; rank-based and related 
measures of association such as 
we have used are recommended instead.
If $f$ is a strictly monotonic, increasing transformation, then the Spearman
(or Kendall) correlation of $f(X_i(t))$ and $f(X_j(t))$ is the same as the
Spearman (or Kendall) correlation of $X_i(t)$ and $X_j(t)$. However, the Pearson
correlation of $f(X_i(t))$ and $f(X_j(t))$ may differ substantially
from the Pearson correlation of $X_i(t)$ and $X_j(t)$. 
Relatedly, if the marginal distributions of $X_i(t)$ and $X_j(t)$ are
changing through time, it may also cause changes through time in the 
Pearson correlation of these quantities, even while the values
of rank-based correlations such as Spearman and Kendall may be unchanging.
Essentially, in such a case, associations between $X_i(t)$ and $X_j(t)$
would be unchanging through time even through Pearson correlation
appears to inducate a change in association, which is instead a change 
in marginal distributions.
Perhaps unfortunately, the Pearson correlation,
which is the most commonly used measure of association of two variables,
depends on both marginal information and association information.
It is possible that recently observed changes in environmental
(e.g., @Black2018) and biological (e.g., @Koenig2016) synchrony 
are, at least in part, just another feature of 
already-studied changes in distributional characteristics of  
variables, rather than reflecting true changes in association.
Rank-based methods are linked to a well-developed suite of statistical
approaches related to the *copula* concept of statistics. Copulas and related ideas
can be used to formally separate the information in the joint 
distribution of $X_i$ and $X_j$ into the information that pertains
to the marginal distributions of these two variables, and the information
that pertains solely to their association. 
@Ghosh_copula argued generally for the potential of copulas and related ideas
to improve our understanding of ecology 
(see also @Popovic2018; @Valpine2014; @Anderson2018).
Rank-based methods such as the Spearman and partial Spearman correlations
which we have used are closely related to copula approaches, and @Ghosh_copula,
@Genest2007 and others have recommended copula and rank approaches when the 
goal is to understand associations between variables, and how such 
associations may be changing.

>>>>

The stochastic quantities $s_e(t)$ and $\tilde{f}_a(t)$ were generated 
to have the same overall association (Spearman's $\rho$ equal to $0.875$,
up to sampling variance), but different degrees and types of 
asymmetry of tail association in different simulations, as follows.
For a simulation of length $T$, we used methods that can 
generate $T$ independent pairs $(u(t),v(t))$ for $t=1,\ldots,T$ with
$\rho=0.875$ (up to sampling variance), but with either extreme left-,
moderate left-, moderate right-, or extreme right-tail association 
(e.g., Fig. \ref{fig_intro_cop}). The methods we used generated
$(u(t),v(t))$ such that $u(t)$ was uniformly distributed on $(0,1)$,
and likewise for $v(t)$. We then applied a monotonic transformation
to the $u(t)$ so that the resulting numbers were distributed according to
a beta distribution with parameters
$\alpha=3$ and $\beta=2$, 
and we used those values for the $s_e(t)$. We also applied a (different)
monotonic transformation to the $v(t)$ so that the resulting numbers were
distributed according to a gamma distribution with parameters $k=2$
and $\theta=1$, and we used those values for the $\tilde{f}_a(t)$. Because
the $u(t)$ and $v(t)$ were uniformly distributed, the inverse 
cumulative distribution functions of the desired beta and
gamma distributions were the transformations. Because these
were monotonic transformations, the Spearman correlation of 
$s_e(t)$ and $\tilde{f}_a(t)$ was the same as that of 
$u(t)$ and $v(t)$. The quantities $(u(t),v(t))$ for $t=1,\ldots,T$ were
generated as described in Appendix \ref{SM-makenoise}.

@@

We simulated our metapopulation models using environmental noises $\epsilon_i(t)$ (Table \ref{tab_model_summary}), 
independent through time, that were generated to have the same overall associations 
between all patches and in all simulations (Spearman's $\rho$ again `r srho_common`, up to sampling variance), 
but with different degrees and types of asymmetry of tail associations 
between patches in different simulations, as follows. For a simulation of length $T$,
we used methods that can generate $T$ independent $n$-tuples $(u_1(t),\ldots,u_n(t))$
for $t=1,\ldots,T$ with $\rho=$ `r srho_common` between $u_i(t)$ and $u_j(t)$ for all $i \neq j$
(up to sampling variance), but with either extreme left-, moderate left-, moderate right-,
or extreme right-tail association between all pairs of locations (as in Fig. \ref{fig_intro_cop}). 
Pairs $u_i(t)$ and $u_j(t)$ looked like Fig. \ref{fig_intro_cop} and were structured, statistically,
in the same way as the stochastic quantities $u(t)$ and $v(t)$ used for the non-spatial 
simulations described above. For any $i$, $u_i(t)$ was again uniformly distributed on 
the interval $(0,1)$. To obtain the values $\epsilon_i(t)$, the $u_i(t)$ were 
transformed so that the resulting numbers were
normally distributed with mean $0$ and standard deviation $\sigma$ (Table \ref{tab_model_summary}), 
for all $i$. This was 
accomplished with the inverse cumulative distribution function of the normal distribution
with mean $0$ and standard deviation $\sigma$. Because this was a monotonic transformation,
the Spearman correlation between $\epsilon_i(t)$ and $\epsilon_j(t)$ was the same 
as that between $u_i(t)$ and $u_j(t)$; likewise partial Spearman correlations were the same. 
The quantities $(u_1(t),\ldots,u_n(t))$
for $t=1,\ldots,T$ were generated as described in Appendix \ref{SM-makenoise}.

@@

Because the deterministic, one-patch versions of our models showed undercompensatory
dynamics for $r<r_c$ (and therefore for $r$ on the left end of the range $(r_{min},r_{bf})$),
and showed overcompensatory dynamics for $r>r_c$, Figs \ref{fig_extrisk_allmodel_localdisp}
and \ref{SM-fig_extrisk_allmodel_globaldisp} suggest the hypothesis that the
opposite effects of tail association on extinction risk correspond to the domains
in which models exhibit under- versus overcompensatory dynamics. This was 
approximately, but not exactly, true. Careful inspection of Figs 
\ref{SM-fig_risk_local_disp_verhulst_r_vary_scl_0.6} - \ref{SM-fig_risk_local_disp_abrewer_r_vary_scl_0.6} and Figs \ref{SM-fig_risk_global_disp_verhulst_r_vary_scl_0.6} - \ref{SM-fig_risk_global_disp_abrewer_r_vary_scl_0.6} 
and the $r_c$ values of Table \ref{tab_model_summary} reveals that, for the Velhurst and Austin-Brewer models,
the switch occurs at a value of $r$ less than $r_c$ for these models. For the example
case of $d=0$, plots of extinction risk versus $r$ (Fig. \ref{fig_extrisk_d0}) seem to indicate that,
for the Ricker, Hassell, and Pennycuick models, left-tail association accentuates and
right-tail association mitigates extinction risk for undercompensatory dynamics ($r<r_c$),
whereas the reverse occurs for overcompensatory dynamics ($r>r_c$). Fig. \ref{fig_extrisk_d0}
reveals a similar reversal for the Maynard Smith, Velhurst and Austin-Brewer models; but 
it occurs at a value of $r$ close to but distinguishable from $r_c$.

@@

<!--linking back to ECEs-->
Extensive evidence now exists that ECEs have become more frequent and more intense
over the last several decades, and substantial progress has recently been made so that
some event categories and even specific events can now be attributed to
human-induced climate change [@ummenhofer2017extreme]. We here provide a fuller 
explanation of how such changes may modify the asymmetry of tail association of 
weather variables measured in different places, and thereby influence metapopulation
extinction risk through the mechanisms we explored in this paper. 
If $X_i(t)$ denotes the value of an environmental variable (e.g., a summer maximum 
temperature, or winter minimum temperature, spring rainfall, etc.) 
in location $i$ in year $t$,
@ummenhofer2017extreme provide a diagram of how ECEs may be increasing in frequency 
and intensity due to changes in the distribution of $X_i(t)$ such as changes
in the mean, variance, or skewness of this distribution. Although changes in 
the correlation between $X_i(t)$ and $X_j(t)$ are increasingly documented, 
and may also be a consequence of climate change (e.g., @Black2018),
we are unaware of any studies examining the extent to which tail associations
between $X_i(t)$ and $X_j(t)$, and asymmetries of tail association, may be 
changing. Documented changes in correlation between $X_i(t)$ and $X_j(t)$
may be due partly or wholly to changes in tail associations. If ECEs are becoming
more spatially extensive, as seems likely given that they are becoming
more intense [@ummenhofer2017extreme], then the conditional probability that both of
$X_i(t)$ and $X_j(t)$ are influenced by an ECE given that one of them is so
influenced should be increasing. This should contribute to increased association
of the two variables in one of their tails (which tail depends on the type of 
ECE and what the variable $X$ is). A study of changes through time in 
tail associations of weather variables would be straightforward and 
potentially important given the mechanisms we have explored.
<!--***DAN: Shyamolina, another aspect of an exploration you may do on the 
the geography of tail association may be to look at how tail association 
may be changing. This relates to ECEs, so may help better justify the work
to a funder. You had proposed to look at geography of tail association in 
NDVI, and that makes good sense. You could also look at how tail association
is changing in NDVI (the records are about 35 years long, so you could just
about do this). Even better (because data are better), you could get weather
station records (lots of these can be downloaded, over 1000 stations
have been operating for over a century continuously in North America) and 
look at how tail associations are changing in those data.-->

<!--copulas and rank methods-->
It is known that the distributions $X_i(t)$ are changing, both in
mean and variance, and possibly in other ways, depending on what
environmental variable $X$ represents.
Studying associations between the values of an environmental variable in different
locations, $X_i(t)$ and $X_j(t)$, using standard methods such as Pearson correlation
may be complicated by these changes; rank-based and related 
measures of association such as 
we have used are recommended instead.
If $f$ is a strictly monotonic, increasing transformation, then the Spearman
(or Kendall) correlation of $f(X_i(t))$ and $f(X_j(t))$ is the same as the
Spearman (or Kendall) correlation of $X_i(t)$ and $X_j(t)$. However, the Pearson
correlation of $f(X_i(t))$ and $f(X_j(t))$ may differ substantially
from the Pearson correlation of $X_i(t)$ and $X_j(t)$. 
Relatedly, if the marginal distributions of $X_i(t)$ and $X_j(t)$ are
changing through time, it may also cause changes through time in the 
Pearson correlation of these quantities, even while the values
of rank-based correlations such as Spearman and Kendall may be unchanging.
Essentially, in such a case, associations between $X_i(t)$ and $X_j(t)$
would be unchanging through time even though Pearson correlation
appears to indicate a change in association, which is instead a change 
in marginal distributions.
Perhaps unfortunately, the Pearson correlation,
which is the most commonly used measure of association of two variables,
depends on both marginal information and association information.
It is possible that recently observed changes in environmental
(e.g., @Black2018) and biological (e.g., @Koenig2016) synchrony 
are, at least in part, just another feature of 
already-studied changes in distributional characteristics of 
variables, rather than reflecting true changes in association.
Rank-based methods are linked to a well-developed suite of statistical
approaches related to the *copula* concept of statistics. Copulas and related ideas
can be used to formally separate the information in the joint 
distribution of $X_i$ and $X_j$ into the information that pertains
to the marginal distributions of these two variables, and the information
that pertains solely to their association. 
@Ghosh_copula argued generally for the potential of copulas and related ideas
to improve our understanding of ecology 
(see also @Popovic2018; @Valpine2014; @Anderson2018).
Rank-based methods such as the Spearman and partial Spearman correlations
which we have used are closely related to copula approaches, and @Ghosh_copula,
@Genest2007 and others have recommended copula and rank approaches when the 
goal is to understand associations between variables, and how such 
associations may be changing.

### 2019-11-05 SuppMat.Rmd
edits

<!--# Algorithm to generate extreme tail association  \label{extr_tail_dep}-->
Bivarite environmental noise with extreme left-tail association (as in Fig. 
\ref{fig_intro_cop}A) was generated using an algorithm originally presented 
by @Ghosh_copula (their Appendix S10), presented again here for convenience.
If $T$ time steps of noise were desired, step 1 was to generate a $T$ by $2$
matrix, $M$, of independent random numbers, each standard-normally distributed. 
Then, each row of $M$ was identified for modification randomly and
independently, with a 50\% chance. For those rows identified, every element
of the row was replaced by the negative of the absolute value of the first element
of the row, including the first element itself. For the other rows, every 
element in the row was replaced by its own absolute value. For noise
with extreme right-tail association (as in Fig. 
\ref{fig_intro_cop}D), take the negative of the entire matrix. For
both types of noise, every entry was transformed by the cummulative distribution
function of a standard normal distribution, to make each column uniformly
distributed.

<!--Spearman correlation of the extreme-tail-associated noise-->
The Spearman correlation of the extreme left- or right-tail associated 
noise was $0.875$, up to sampling variation.
<!--***DAN: fill this in-->

<!--moderate tail association noise-->
Bivarite environmental noise with moderate left-tail association (as in Fig. 
\ref{fig_intro_cop}B) was generated using a 2-dimensional Clayton copula. The statistical
theory of copulas has been introduced systematically in several texts 
(e.g., @nelsen2006_copula; @joe2014_dependence), and was briefly 
introduced by @Genest2007. Short, conceptual introductions aimed at ecological
audiences were provided by sections within the papers of @Anderson2018 and  
@Ghosh_copula. A 2-dimensional Clayton copula is equivalent to a particular bivariate 
probability density function with marginal distributions which are 
uniform on the interval $(0,1)$. The points on Fig. 
\ref{fig_intro_cop}B were independent draws from the associated random variable.
The functions `claytonCopula` and `rCopula` from the `copula` package for the 
R programming language were used. The parameter of the Clayton family was
chosen so that the Spearman correlation was $0.875$, the same as that of the
extreme left-tail associated noise. This was accomplished with the `iRho`
function of the `copula` package. Noise with moderate right-tail
association was generated in a similar way but using the 2-dimensional 
survival Clayton copula.

<!--Multivariate extreme tail association-->
Multivariate noise with extreme left-tail association was generated using a
very simlar algorithm to the bivariate case,
as follows. If $T$ time steps of noise of dimension $n$ were desired,
step 1 was to generate a $T$ by $n$
matrix, $M$, of independent random numbers, each standard-normally distributed.
Each row of $M$ was identified for modification randomly and
independently, with a 50\% chance. For those rows identified, every element
of the row was replaced by the negative of the absolute value of the first element
of the row, including the first element itself. For the other rows, every 
element in the row was replaced by its own absolute value. For noise
with extreme right-tail association, take the negative of the entire matrix. For
both types of noise, every entry was transformed by the cummulative distribution
function of a standard normal distribution, to make each column uniformly
distributed. 
Multivariate moderately tail associated noise of dimension $n$
was generated using $n$-dimensional Clayton or survival Clayton copulas.

@@

**The Austin-Brewer model.** The Austin-Brewer model 
[@austin1971world; @cohen1995unexpected] was originally used 
to model the growth of human populations. The model is $P(t+1) = F(P(t),r,K,s)= P(t)[1+r(K-P(t)) \left\{ 1-\exp (-sP(t))\right\}]$, where
$r, K, s$ are positive parameters. $r$ and $K$ represent the growth rate and carrying 
capacity, respectively.
the usual extinction equilibrium is $P_e=0$. The 
positive equilibrium is $P_e=P^\star=1-rK\left\{ 1-\exp (-sK)\right\}$. $P^\star$
is stable when $r<r_{bf}$ and unstable beyond than bifurcation point, where 
$r_{bf} = \frac{2}{K[1-\exp (-sK)]}$. The model shows undercompensatory dynamics when
$0<r<r_c$, where $r_c = \frac{1}{K[1-\exp (-sK)]}$. It shows 
overcompensatory dynamics when $r_c<r<r_{bf}$.

>>>>

<!--# Algorithm to generate extreme tail association  \label{extr_tail_dep}-->
Bivarite environmental noise with extreme left-tail association (as in Fig. 
\ref{MT-fig_intro_cop}A) was generated using an algorithm originally presented 
by @Ghosh_copula (their Appendix S10), presented again here for convenience.
If $T$ time steps of noise were desired, step 1 was to generate a $T$ by $2$
matrix, $M$, of independent random numbers, each standard-normally distributed. 
Then, each row of $M$ was identified for modification randomly and
independently, with a 50\% chance. For those rows identified, every element
of the row was replaced by the negative of the absolute value of the first element
of the row, including the first element itself. For the other rows, every 
element in the row was replaced by its own absolute value. For noise
with extreme right-tail association (as in Fig. 
\ref{MT-fig_intro_cop}D), take the negative of the entire matrix. For
both types of noise, every entry was transformed by the cumulative distribution
function of a standard normal distribution, to make each column uniformly
distributed.

<!--Spearman correlation of the extreme-tail-associated noise-->
The Spearman correlation of the extreme left- or right-tail associated 
noise was `r srho_common`, up to sampling variation.
<!--moderate tail association noise-->
Bivarite environmental noise with moderate left-tail association (as in Fig. \ref{MT-fig_intro_cop}B) was 
generated using a 2-dimensional Clayton copula. The statistical
theory of copulas has been introduced systematically in several texts 
(e.g., @nelsen2006_copula; @joe2014_dependence), and was briefly 
introduced by @Genest2007. Short, conceptual introductions aimed at ecological
audiences were provided by sections within the papers of @Anderson2018 and @Ghosh_copula. A
2-dimensional Clayton copula is equivalent to a particular bivariate 
probability density function with marginal distributions which are 
uniform on the interval $(0,1)$. The points on Fig. 
\ref{MT-fig_intro_cop}B were independent draws from the associated random variable.
The functions `claytonCopula` and `rCopula` from the `copula` package for the 
R programming language were used. The parameter of the Clayton family was
chosen so that the Spearman correlation was `r srho_common`, the same as that of the
extreme left-tail associated noise. This was accomplished with the `iRho`
function of the `copula` package. Noise with moderate right-tail
association was generated in a similar way but using the 2-dimensional 
survival Clayton copula.

<!--Multivariate extreme tail association-->
Multivariate noise with extreme left-tail association was generated using a
very simlar algorithm to the bivariate case,
as follows. If $T$ time steps of noise of dimension $n$ were desired,
step 1 was to generate a $T$ by $n$
matrix, $M$, of independent random numbers, each standard-normally distributed.
Each row of $M$ was identified for modification randomly and
independently, with a 50\% chance. For those rows identified, every element
of the row was replaced by the negative of the absolute value of the first element
of the row, including the first element itself. For the other rows, every 
element in the row was replaced by its own absolute value. For noise
with extreme right-tail association, take the negative of the entire matrix. For
both types of noise, every entry was transformed by the cumulative distribution
function of a standard normal distribution, to make each column uniformly
distributed. 
Multivariate moderately tail associated noise of dimension $n$
was generated using $n$-dimensional Clayton or survival Clayton copulas.

@@

**The Austin-Brewer model.** The Austin-Brewer model 
[@austin1971world; @cohen1995unexpected] was originally used 
to model the growth of human populations. The model is $P(t+1) = F(P(t),r,K,s)= P(t)[1+r(K-P(t)) \left\{ 1-\exp (-sP(t))\right\}]$, where
$r, K, s$ are positive parameters. $r$ and $K$ represent the growth rate and carrying 
capacity, respectively.
the usual extinction equilibrium is $P_e=0$. The 
positive equilibrium is $P_e=P^\star=1-rK\left\{ 1-\exp (-sK)\right\}$. $P^\star$
is stable when $r<r_{bf}$ and unstable beyond the bifurcation point, where 
$r_{bf} = \frac{2}{K[1-\exp (-sK)]}$. The model shows undercompensatory dynamics when
$0<r<r_c$, where $r_c = \frac{1}{K[1-\exp (-sK)]}$. It shows 
overcompensatory dynamics when $r_c<r<r_{bf}$.
